---
title: "Project1:Chess Tournament Data"
author: "Ummay Rukiya"
date: "2026-09-25"
output: html_document
---

## Overview

For this project, I will use R to extract player information from a text file containing chess tournament results. The final dataset will include each player’s name, state, total points, pre-tournament rating, and average pre-tournament rating of their opponents. I will then export the completed dataframe as a CSV file.

## Loading the Tournament Data

I loaded the original text file into R as separate lines. I will use the repeated structure of the file to identify the player records and extract the required information.

```{r load-data}
tournament_lines <- readLines(
  "tournamentinfo.txt",
  warn = FALSE
)

length(tournament_lines)
head(tournament_lines, 10)
```

## Identifying Player Records

Each player’s information appears across two lines. The first line contains the player’s number, name, points, and opponents. The second line contains the player’s state and pre-tournament rating. I identified these two types of lines using their repeated patterns.

```{r identify-records}
player_line_numbers <- grep(
  "^\\s*[0-9]+\\s*\\|",
  tournament_lines
)

rating_line_numbers <- grep(
  "^\\s*[A-Z]{2}\\s*\\|",
  tournament_lines
)

player_lines <- tournament_lines[player_line_numbers]
rating_lines <- tournament_lines[rating_line_numbers]

length(player_lines)
length(rating_lines)
```

## Extracting Player Information

I separated each record at the vertical bars in the text file. From these fields, I extracted each player’s tournament number, name, total points, state, and pre-tournament rating.

```{r extract-player-information}
split_record <- function(record) {
  trimws(strsplit(record, "|", fixed = TRUE)[[1]])
}

player_fields <- lapply(player_lines, split_record)
rating_fields <- lapply(rating_lines, split_record)

pair_number <- as.integer(
  sapply(player_fields, function(record) record[1])
)

player_name <- sapply(
  player_fields,
  function(record) record[2]
)

total_points <- as.numeric(
  sapply(player_fields, function(record) record[3])
)

player_state <- sapply(
  rating_fields,
  function(record) record[1]
)

rating_text <- sapply(
  rating_fields,
  function(record) record[2]
)

pre_rating <- as.integer(
  sub(".*R:\\s*([0-9]+).*", "\\1", rating_text)
)

player_information <- data.frame(
  pair_number,
  player_name,
  player_state,
  total_points,
  pre_rating,
  stringsAsFactors = FALSE
)

head(player_information)
```

## Extracting Opponent Numbers

Each completed round contains a result—win, loss, or draw—followed by the opponent’s player number. Entries such as byes, half-point byes, and unplayed rounds do not identify an opponent, so I excluded them.

```{r extract-opponents}
round_results <- lapply(
  player_fields,
  function(record) record[4:10]
)

extract_opponents <- function(rounds) {
  
  completed_games <- grepl(
    "^[WLD]\\s*[0-9]+",
    rounds
  )
  
  opponent_ids <- sub(
    "^[WLD]\\s*([0-9]+).*$",
    "\\1",
    rounds[completed_games]
  )
  
  as.integer(opponent_ids)
}

opponent_numbers <- lapply(
  round_results,
  extract_opponents
)

opponent_numbers[[1]]
```

## Calculating Average Opponent Ratings

I matched each opponent number with that opponent’s pre-tournament rating. I then calculated the mean of those ratings for every player and rounded the result to the nearest whole number, following the assignment example.

```{r calculate-opponent-averages}
average_opponent_pre_rating <- sapply(
  opponent_numbers,
  function(opponent_ids) {
    
    opponent_ratings <- pre_rating[
      match(opponent_ids, pair_number)
    ]
    
    round(mean(opponent_ratings))
  }
)

chess_results <- data.frame(
  player_name,
  player_state,
  total_points,
  pre_rating,
  average_opponent_pre_rating,
  stringsAsFactors = FALSE
)

head(chess_results)
```

## Final Tournament Results

The final dataframe contains the five required fields for all 64 players. I checked the number of rows and searched for missing values before exporting the results.

```{r validate-results}
nrow(chess_results)
colSums(is.na(chess_results))

knitr::kable(
  chess_results,
  caption = "Chess Tournament Player Results"
)
```

## Exporting the CSV File

I exported the completed dataframe as a CSV file so it can be used in another program or imported into a database.

```{r export-csv}
write.csv(
  chess_results,
  "chess_tournament_results.csv",
  row.names = FALSE
)

file.exists("chess_tournament_results.csv")
```

## Conclusion

I successfully extracted the required information for all 64 chess players from the original tournament text file. I used each player’s round results to identify their opponents and matched those opponent numbers with their pre-tournament ratings. The final dataframe contains each player’s name, state, total points, pre-rating, and average opponent pre-rating. I also exported the completed results as a CSV file that can be imported into a database or used for further analysis.

## Generative AI Use

I used ChatGPT to help develop parts of the R code, understand how to extract information from the structured text file, and check the final calculations.

OpenAI. (2026). *ChatGPT (GPT-5)* [Large language model]. https://chatgpt.com. Accessed September 26, 2026.














